---
title: "equilibrium-and-R0"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{equilibrium-and-R0}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(facilityepimath)
```
The functions in this package analyze compartmental differential equation models, where the compartments are states of individuals who are co-located in a healthcare facility, for example, patients in a hospital or patients/residents of long-term care facility.

Our example here is a hospital model with four compartments: two compartments $C_1$ and $C_2$ representing patients who are colonized with an infectious organism, and two compartments $S_1$ and $S_2$ representing patients who are susceptible to acquiring colonization with that same organism. The two colonized and susceptible states are distinguished by having potentially different infectivity to other patients and vulnerability to acquisition from other patients, respectively.

A system of differential equations with those 4 compartments may take the following general form:

\[ \frac{dS_1}{dt} = -(s_{21}+(a_{11}+a_{21})\alpha+\omega_1+h(t))S_1 + s_{12}S_2 + r_{11}C_1 + r_{12}C_2\]

\[\frac{dS_2}{dt} = s_{21}S_1 - (s_{12}+(a_{12}+a_{22})\alpha+\omega_2+h(t))S_2 + r_{21}C_1 + r_{22}C_2 \]

\[\frac{dC_1}{dt} = a_{11}\alpha S_1 + a_{12}\alpha S_2 - (c_{21}+r_{11}+r_{21}+\omega_3+h(t))C_1 + c_{12}C_2 \]

\[\frac{dC_2}{dt} = a_{21}\alpha S_1 + a_{22}\alpha S_2 + c_{21}C_1 - (c_{12}+r_{12}+r_{22}+\omega_4+h(t))C_2 \]

The acquisition rate $\alpha$ appearing in each equation, and governing the transition rates between the S compartments and the C compartments, is assumed to depend on the number of colonized patients in the facility, as follows:

\[ \alpha = \beta_1 C_1 + \beta_2 C_2 \]

## Facility Basic Reproduction Number $R_0$

We will demonstrate how to calculate the basic reproduction number $R_0$ of this system using the `facilityR0` function. The following components of the system are required as inputs to the function call below.

### The `S` matrix; pre-invasion susceptible state transitions

A matrix `S` governing the transitions between, and out of, the states $S_1$ and $S_2$ in the absence of any colonized patients "invading" the facility (i.e., all state transitions except for acquiring colonization):

\[S = \left(
\begin{matrix}
    -s_{21}-\omega_1 & s_{12} \\
    s_{21} & -s_{12}-\omega_2
\end{matrix}\right)
\]

```{r}
s21 <- 1; s12 <- 2
omega1 <- omega2 <- 0.1
S <- rbind(c(-s21-omega1, s12), c(s21, -s12-omega2))
```
The $s_{ij}$ rates can be used to model transitions between different patient states that might alter susceptibility to acquiring the modeled organism, for example risky treatment procedures, drug exposures, or protective measures. The $\omega_i$ rates model removal from the facility via discharge or death.

### The `C` matrix; colonized state transitions

Next, we require a matrix `C` governing the transitions between, and out of, the states $C_1$ and $C_2$:

\[C = \left(
\begin{matrix}
    -c_{21}-r_{11}-r_{21}-\omega_3 & c_{12} \\
    c_{21} & -c_{12}-r_{12}-r_{22}-\omega_4
\end{matrix}\right)
\]
```{r}
c21 <- 0.1; c12 <- 0
r11 <- r22 <- 0.1; r12 <- r21 <- 0
omega3 <- omega4 <- 0.1
C <- rbind(c(-c21-r11-r21-omega3, c12), c(c21, -c12-r12-r22-omega4))
```
The $c_{ij}$ rates can be used to model transitions between different colonized patient states that might be observable in data and/or alter the patient's infectivity to other patients, for example clinical infection, detection status, or placement under protective measures. The $r_{ij}$ rates govern clearance of colonization (thus, transitions back to one of the susceptible states), while the $\omega_i$ rates model removal from the facility via discharge or death.

### The `A` matrix; susceptible-to-colonized state transitions and relative susceptibility
The next input requirement is a matrix `A` describing the S-to-C state transitions when an acquisition occurs:

\[A = \left(
\begin{matrix}
    a_{11} & a_{12} \\
    a_{21} & a_{22}
\end{matrix}\right)
\]
```{r}
a11 <- 1; a22 <- 2; a12 <- a21 <- 0
A <- rbind(c(a11,a12),c(a21,a22))
```
The $a_ij$ rates describe which of the two colonized states can be entered from each of the two susceptible states at the moment of acquiring colonization; in the above example, state $S_1$ patients move only to state $C_1$ and state $S_2$ patients move only to state $C_2$. The $a_ij$ values also describe the \emph{relative} susceptibility of the two $S$ states; in the above example, state $S_2$ patients are twice as susceptible as state $S_1$ patients.

### The `transm` vector: transmission rates
The next required input is a vector `transm` containing the $\beta$ coefficients (transmission rates from each colonized compartment) appearing in the $\alpha$ equation above: $(\beta_1,\beta_2)$
```{r}
beta1 <- 0.2; beta2 <- 0.3
transm <- c(beta1, beta2)
```
The $\beta_j$ values describe the transmissibility of each colonized state; in this example, state $C_2$ patients are 50% more transmissible than state $C_1$ patients. When the levels of colonization in the facility are $C_1$ and $C_2$, the acquisition rate of an $S_i$ patient is $a_{ii}(\beta_1C_1+\beta_2C_2)$

### The `initS` vector: susceptible state admission distribution
A vector `initS` containing the admission state probabilities for the susceptible compartments only (i.e., the pre-invasion system before a colonized patient is introduced): $(\theta_1,1-\theta_1)$

```{r}
theta1 <- 0.9; theta2 <- 1 - theta1
initS <- c(theta1, theta2)
```
The `initS` vector should sum to 1.

### `facilityR0` function call with $h(t)=0$

The time-of-stay-dependent removal rate $h(t)$ is the remaining component in the master equations above. When $h(t)=0$, the above components are sufficient to calculate the facility $R_0$ as in the following example:

```{r}
facilityR0(S,C,A,transm,initS)
```
Note that when $h(t)=0$, the removal rates (discharge and death) are entirely governed by the $\omega_i$ values, which must be set such that patients in any state must be guaranteed to eventually reach a state for which $\omega_i>0$; otherwise, patients can have an infinitely long length of stay in the facility.

### The `mgf` function: moment generating function associated with $h(t) != 0$

When the time-of-stay-dependent removal rate $h(t)\neq0$, there is one additional argument required as input to the `facilityR0` function: a function `mgf(x,deriv)` that is the moment-generating function (and its derivatives) of the distribution for which $h(t)$ is the hazard function. This is the length of stay distribution when the state-dependent removal rates $\omega$ are all zero, as in the following example.
```{r}
omega1 <- omega2 <- omega3 <- omega4 <- 0
S <- rbind(c(-s21-omega1, s12), c(s21, -s12-omega2))
C <- rbind(c(-c21-r11-r21-omega3, c12), c(c21, -c12-r12-r22-omega4))
```
The length of stay distribution can be any statistical distribution with non-negative range, as long as the moment generating function (mgf) and its derivatives can be evaluated, as this calculation is required within the $R_0$ formula. We currently provide three mgf functions in this package: one for the exponential distribution, `MGFexponential`, one for the gamma distribution `MGFgamma`, and one for a mixed gamma distribution `MGFmixedgamma`; the latter distribution can employ a weighted mixture of any number of different gamma distributions.

The `mgf(x, deriv)` function to be passed to `facilityR0()` must be defined as in the following example, which uses a gamma distribution.

```{r}
mgf <- function(x, deriv=0) MGFgamma(x, rate=0.1, shape=3.1, deriv)
```

### `facilityR0` function call with $h(t)\neq0$

The call to `facilityR0()` is the same as above with the additional argument `mgf` (which defaulted to `NULL` when not provided):

```{r}
facilityR0(S,C,A,transm,initS,mgf)
```

## Facility model equilibrium

We will demonstrate how to calculate the equilibrium of the full system of equations, with a given set of initial conditions, i.e. distribution of patient states at admission, using the `facilityeq` function. The following components of the system are required as inputs to the function call below.

